---
title: "LIANA_data_preparation"
output: html_document
date: "2024-10-22"
---

In this markdown, I am taking the outputs from the `RPM_cellchat_individual_TP_analysis.Rmd` and `RPM_DEA_decoupleR.Rmd` markdowns and preparing them for the LIANA+ workflow. I am using the RPM SCLC data as an example but the code is the same for the datasets. Also, I am only focusing on TP7 and TP11 in this dataset since this is when both the NE/epithelial and non-NE/mesenchymal populations are present.

# Load necessary packages
```{r}
library(tidyverse)
library(OmnipathR)
library(reticulate)
```

# Read in data
```{r}
# Function to load and format data
load_and_format_data <- function(matrix_file, var_file, obs_file) {
  # Import sparse matrix
  scipy_sparse <- import("scipy.sparse")
  tp_matrix <- scipy_sparse$load_npz(matrix_file)
  
  # Load var and obs files
  var_data <- read.csv(var_file, header = TRUE)
  obs_data <- read.csv(obs_file, header = TRUE)
  
  # Set row names for obs_data
  rownames(obs_data) <- obs_data$ID
  
  # Format the matrix
  df <- as.data.frame(as.matrix(tp_matrix))
  df <- t(df)
  colnames(df) <- obs_data[, 1]
  rownames(df) <- var_data[, 1]
  
  return(df)
}

# Define file paths for the different datasets
matrix_files <- c(
  "/home/dlopez7/SCLC/RPM/matrix_data/TP7_norm_data.npz",
  "/home/dlopez7/SCLC/RPM/matrix_data/TP11_norm_data.npz"
)

var_files <- c(
  "/home/dlopez7/SCLC/RPM/matrix_data/AD_norm_TP7_var.csv",
  "/home/dlopez7/SCLC/RPM/matrix_data/AD_norm_count_TP11.csv"
)

obs_files <- c(
  "/home/dlopez7/SCLC/RPM/matrix_data/AD_norm_TP7_obs.csv",
  "/home/dlopez7/SCLC/RPM/matrix_data/AD_norm_count_TP11_obs.csv"
)

# Load and format the data for each dataset
df_7 <- load_and_format_data(matrix_files[1], var_files[1], obs_files[1])
df_11 <- load_and_format_data(matrix_files[2], var_files[2], obs_files[2])
```

# L-R interaction preparation
```{r}
# Read in the LR interaction list from the merged cellchat object
dfLR.net <- readRDS("/home/dlopez7/SCLC/RPM/rds/mouse_cellchat_LR_int.rds")
```

```{r}
# Extract TP4 LR interactions
df7.net <- dfLR.net$TP7
df11.net <- dfLR.net$TP11
```

Normalize the cellchat interaction scores from 0-1 since some of the values are really low.
```{r}
# Function to normalize LR scores between [0,1]
normalize_LR <- function(x){
  LR_strength <- x$prob
  x$norm_str <- rescale(LR_strength, to = c(0,1))
}

# Normalize LR scores for each timepoint
df7.net$norm_str <- normalize_LR(df7.net)
df11.net$norm_str <- normalize_LR(df11.net)
```

This next part is only needed for mouse data since OmniPath does not contain L-R complexes for mouse genes.
```{r}
# Function to split up receptor complexes
split_LRs <- function(x){
  x %>%
    separate_rows(receptor, sep = '_') %>%
    mutate(receptor = str_to_title(receptor)) %>%
    distinct(source, target, ligand, receptor, .keep_all = TRUE)
}

df7.net <- split_LRs(df7.net)
df11.net <- split_LRs(df11.net)
```

Save the different LR dataframes

# OmniPath interactions
```{r}
# protein-protein interactions
# ppi <- import_omnipath_interactions(organism = 10090)
ppi_with_KE_PE <- import_omnipath_interactions(datasets= c('omnipath', 'pathwayextra', 'kinaseextra'), organism = 10090)

# transcriptional regulation interactions
TF_regulons <- import_transcriptional_interactions(organism = 10090)
```

Save the ppi_with_KE_PE and TF_regulons dataframes

# TF activity results
```{r}
# Load TF results
TP7_A2_TFs <- readRDS("/home/dlopez7/SCLC/RPM/rds/TP7_TFs_activity_A2.rds")
TP7_AN_TFs <- readRDS("/home/dlopez7/SCLC/RPM/rds/TP7_TFs_activity_AN.rds")
TP7_PY_TFs <- readRDS("/home/dlopez7/SCLC/RPM/rds/TP7_TFs_activity_PY.rds")

TP11_A2_TFs <- readRDS("/home/dlopez7/SCLC/RPM/rds/TP11_TFs_activity_A2.rds")
TP11_PY_TFs <- readRDS("/home/dlopez7/SCLC/RPM/rds/TP11_TFs_activity_PY.rds")
```

```{r}
# Function to filter TF activity df
TF_filter <- function(TF_df, norm_countDF){
  TF_df %>%
    filter(source %in% toupper(rownames(norm_countDF))) %>%
    filter(p_value < 0.05 & score > 0) %>%
    arrange(desc(score))
}

active_A2_TFs_TP7 <- TF_filter(TP7_A2_TFs, df_7)
active_AN_TFs_TP7 <- TF_filter(TP7_AN_TFs, df_7)
active_PY_TFs_TP7 <- TF_filter(TP7_PY_TFs, df_7)

active_A2_TFs_TP11 <- TF_filter(TP11_A2_TFs, df_11)
active_PY_TFs_TP11 <- TF_filter(TP11_PY_TFs, df_11)
```

Save the TFs

# Differential Expression markers
```{r}
# Load DEA results
TP7_markers_A2 <- read_rds('/home/dlopez7/SCLC/RPM/rds/TP7_markers_A2.rds')
TP7_markers_AN <- read_rds('/home/dlopez7/SCLC/RPM/rds/TP7_markers_AN.rds')
TP7_markers_PY <- read_rds('/home/dlopez7/SCLC/RPM/rds/TP7_markers_PY.rds')

TP11_markers_A2 <- read_rds('/home/dlopez7/SCLC/RPM/rds/TP11_markers_A2.rds')
TP11_markers_PY <- read_rds('/home/dlopez7/SCLC/RPM/rds/TP11_markers_PY.rds')
```

```{r}
# Filter for significance
TP7_markers_A2 <- TP7_markers_A2 %>% filter(p_val_adj < 0.05)
TP7_markers_AN <- TP7_markers_AN %>% filter(p_val_adj < 0.05)
TP7_markers_PY <- TP7_markers_PY %>% filter(p_val_adj < 0.05)

TP11_markers_A2 <- TP11_markers_A2 %>% filter(p_val_adj < 0.5)
TP11_markers_PY <- TP11_markers_PY %>% filter(p_val_adj < 0.5)
```

Save the filtered DEA results