---
title: "RPM CellChat Individual Analysis"
output: html_document
date: "2024-10-21"
---

In this markdown, I am performing the analysis of an individual timepoint within the SCLC RPM dataset. The same approach is applied to the other timepoints before the cellchat objects together and performing a stratified comparison. For this markdown, I am using TP7, but the same code is used for the other timepoints.

# Load required libraries
```{r}
library(CellChat)
library(patchwork)
library(tidyverse)
library(reticulate)
options(stringsAsFactors = FALSE)
```

# Read in the data files
```{r}
# Import sparse matrix
scipy_sparse = import("scipy.sparse")
tp7_csr_matrix = scipy_sparse$load_npz("/home/dlopez7/SCLC/RPM/matrix_data/TP7_norm_data.npz")
```

```{r}
# Load in var and obs files from the anndata object
TP7_var <- read.csv("/home/dlopez7/SCLC/RPM/matrix_data/AD_norm_TP7_var.csv", header = TRUE)

TP7_obs <- read.csv("/home/dlopez7/SCLC/RPM/matrix_data/AD_norm_TP7_obs.csv", header = TRUE)
rownames(TP7_obs) <- TP7_obs$ID
```

```{r}
# Format the matrix
df_7 <- as.data.frame(as.matrix(tp7_csr_matrix))
df_7 <- t(df_7)
colnames(df_7) <- TP7_obs[,1]
rownames(df_7) <- TP7_var[,1]
```

# Create CellChat object
```{r}
norm7_cellchat <- createCellChat(object = df_7, 
                                 meta = TP7_obs, 
                                 group.by = "ParetoTI_specialists")
```

# Set the ligand-receptor interaction database
```{r}
CellChatDB <- CellChatDB.mouse
showDatabaseCategory(CellChatDB)

# show structure of the database
dplyr::glimpse(CellChatDB$interaction)

# use all CellChatDB for cell-cell communication analysis
CellChatDB.use <- CellChatDB

# set the used database in the object
norm7_cellchat@DB <- CellChatDB.use
```

# Preprocessing the expression data for cell-cell communication analysis
```{r}
# subset the sexpression data of signaling genes to save on computation cost
norm7_cellchat <- subsetData(norm7_cellchat)

norm7_cellchat <- identifyOverExpressedGenes(norm7_cellchat)
norm7_cellchat <- identifyOverExpressedInteractions(norm7_cellchat)
```

```{r}
# Inference of cecll-cell communication network
norm7_cellchat <- computeCommunProb(norm7_cellchat)
```

```{r}
# Filter out cell-cell communication if there are only a few cells in cell groups
norm7_cellchat <- filterCommunication(norm7_cellchat, min.cells = 10)
```

```{r}
# Infer the cell-cell communication at a signaling pathway level
norm7_cellchat <- computeCommunProbPathway(norm7_cellchat)
```

```{r}
# Calculate the aggregaated cell-cell communication network
norm7_cellchat <- aggregateNet(norm7_cellchat)
```

# CellChat System Analysis
```{r}
# Compute network centrality scores
norm7_cellchat <- netAnalysis_computeCentrality(norm7_cellchat, slot.name = "netP")
# Visualize computed scores
# netAnalysis_signalingRole_network(norm7_cellchat, width = 8, height = 2.5, font.size = 10)
```

# Save CellChat object
```{r}
# saveRDS(norm7_cellchat, 'norm_count_TP7.rds')
```

# Subset communications
```{r}
df7net <- subsetCommunication(norm7_cellchat)
# Save df
write_tsv(df7net, 'df7Net.txt')
```

```{r}
sessionInfo()
```

