---
title: "RPM decoupleR"
output: html_document
date: "2024-10-21"
---

In this markdown, I am performing the differential expression analysis using Seurat and the transcription factor activity inference using decoupleR. Here, I am only performing the analysis on the RPM SCLC TP7 data but the procedure is the same for the other timepoints and datasets.

# Load necessary packages
```{r}
library(tidyverse)
library(Seurat)
library(decoupleR)
library(reticulate)
```

# Read in data files
```{r}
# Import sparse matrix
scipy_sparse = import("scipy.sparse")
tp7_csr_matrix = scipy_sparse$load_npz("/home/dlopez7/SCLC/RPM/matrix_data/TP7_raw_counts.npz")
```

```{r}
# Load in var and obs files
TP7_var <- read.csv("/home/dlopez7/SCLC/RPM/matrix_data/AD_norm_TP7_var.csv", header = TRUE)

TP7_obs <- read.csv("/home/dlopez7/SCLC/RPM/matrix_data/AD_norm_TP7_obs.csv", header = TRUE)
rownames(TP7_obs) <- TP7_obs$ID
```

```{r}
# Format the matrix
df_7 <- as.data.frame(as.matrix(tp7_csr_matrix))
df_7 <- t(df_7)
colnames(df_7) <- TP7_obs[,1]
rownames(df_7) <- TP7_var[,1]

# Convert matrix to sparse matrix
df_7_sparse <- as(df_7, "sparseMatrix")
```

# Seurat Workflow
```{r}
# Create seurat object
mRPM_TP7 <- CreateSeuratObject(counts = df_7_sparse, project = "Mouse_RPM_TP7")
```

```{r}
# Extract cell type label from obs df for each cell
ParetoTI <- TP7_obs$ParetoTI_specialists

# Add labels to seurat object
mRPM_TP7 <- AddMetaData(mRPM_TP7, metadata = ParetoTI, col.name = "Specialist")
```

```{r}
# Filter Malat1
mRPM_TP7 <- mRPM_TP7[!grepl("Malat1", rownames(mRPM_TP7)), ]

# Filter Mitochondrial
mRPM_TP7 <- mRPM_TP7[!grepl("^Mt-", rownames(mRPM_TP7)), ]

# Filter Ribosomal
mRPM_TP7 <- mRPM_TP7[!grepl("^Rp", rownames(mRPM_TP7)), ]
```

```{r}
# Normalize the data
mRPM_TP7 <- NormalizeData(mRPM_TP7)

# Find Variable Features
mRPM_TP7 <- FindVariableFeatures(mRPM_TP7, selection.method = "vst", nfeatures = 2000)

# Scale data
all.genes7 <- rownames(mRPM_TP7)
mRPM_TP7 <- ScaleData(mRPM_TP7, features = all.genes7)

# Perform linear dimensional reduction
mRPM_TP7 <- RunPCA(mRPM_TP7, features = VariableFeatures(mRPM_TP7), verbose = F)

# Cluster cells
mRPM_TP7 <- FindNeighbors(mRPM_TP7, dims = 1:20)
mRPM_TP7 <- FindClusters(mRPM_TP7, resolution = 0.5)

# Run UMAP
mRPM_TP7 <- RunUMAP(mRPM_TP7, dims = 1:20)
```

```{r}
# Plot UMAP using specialist label
DimPlot(mRPM_TP7, reduction = "umap", group.by = "Specialist")
```

```{r}
# Set identity of object to specialist subtypes
Idents(mRPM_TP7) <- "Specialist"

# Function to find markers
find_markers <- function(seurat_object, group1, group2, min_pct = 0.1, logfc_threshold = 0.1){
  markers <- FindMarkers(seurat_object, ident.1=group1, ident.2=group2, min.pct=min_pct, logfc.threshold=logfc_threshold) %>% 
    mutate(Gene = rownames(.)) %>% 
    filter(p_val_adj <= 0.05)
  
  return(markers)
}

# Find markers for each subtype
A2_markers <- find_markers(mRPM_TP7, 'A2', c('A/N', 'P/Y'))
AN_markers <- find_markers(mRPM_TP7, 'A/N', c('A2', 'P/Y'))
PY_markers <- find_markers(mRPM_TP7, 'P/Y', c('A2', 'A/N'))
```

```{r}
# Save markers df
# write_tsv(A2_markers, '/path/to/file')
# write_tsv(AN_markers, '/path/to/file')
# write_tsv(PY_markers, '/path/to/file')
```

# DecoupleR Workflow
```{r}
# Get CollecTRI network
net <- get_collectri(organism = "mouse", split_complexes = FALSE)
```

```{r}
# Function to extract log-fold values and format df for ulm method
extract_log2 <- function(marker_df){
  log2FC <- data.frame(marker_df$avg_log2FC)
  rownames(log2FC) <- toupper(rownames(marker_df))
  
  return(log2FC)
}

logFC_A2 <- extract_log2(A2_markers) 
logFC_AN <- extract_log2(AN_markers)
logFC_PY <- extract_log2(PY_markers)
```

```{r}
# Run ULM TF activity inference for each cell type
TP7_acts_A2 <- run_ulm(mat = logFC_A2, network = net, .source = "source", .target = "target", .mor = "mor", minsize = 0) 

TP7_acts_AN <- run_ulm(mat = logFC_AN, network = net, .source = "source", .target = "target", .mor = "mor", minsize = 0)

TP7_acts_PY <- run_ulm(mat = logFC_PY, network = net, .source = "source", .target = "target", .mor = "mor", minsize = 0)
```

```{r}
# Save TF activity results
# write_tsv(TP7_acts_A2, '/path/to/file')
# write_tsv(TP7_acts_AN, '/path/to/file')
# write_tsv(TP7_acts_PY, '/path/to/file')
```

